Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections in the LArIAT experiment

### 2018-05-24 Chapter-2/2-LArTPCs.tex
s

\chapter{Liquid Argon Detectors at the Intensity Frontier}\label{ch:2}
{\raggedleft ``\emph{Beat-up little seagull, on a marble stair} \par}
{\raggedleft \emph{Tryin' to find the ocean, lookin' everywhere.}"\par}
{\raggedleft -- Nina Simone, Baltimore, 1978 -- \par}
\vspace{0.5cm}

In the next few years, LArTPCs will  be the tools to answer some of the burning questions in neutrino physics today.  This section illustrates the operational principles of this detector technology, as well as the scope of the key detectors in the US liquid argon program -- SBN, DUNE and LArIAT.

\section{The Liquid Argon Time Projection Chamber Technology}
\subsection{TPCs, Neutrinos \& Argon}
David Nygren designed the first Time Projection Chamber (TPC) in the late 1970s~\cite{FirstTPC} for the  PEP-4 experiment, a detector  apt to study electron-positron collisions at the PEP storage ring at the SLAC National Accelerator Laboratory.
From the original design  in the seventies -- a cylindrical chamber filled with methane gas -- the TPC detector concept has seen many incarnations, the employment of several different active media and a variety of different particle physics applications, including, but not limited to the study of electron/positron storage rings (e.g. PEP4, TOPAZ, ALEPH and DELPHI), heavy ions collisions in fixed target and collider experiments (e.g. EOS/HISS and ALICE ), dark matter (ArDM), rare decays and capture (e.g. TRIUMP, MuCap),  neutrino detectors and nucleon decay (ICARUS, SBN, DUNE), and neutrino less double beta decay (Next). A nice review of the history of TPCs and working principles is provided in \cite{0034-4885-73-11-116201}.

@@

\section{The Intensity Frontier Program}
\subsection{Prospects for LArTPCs in Neutrino Physics: SBN and DUNE}
The ArgoNeut  experiment \cite{ArgoNeuT-det} initiated the US LArTPC neutrino program. Following the success of this small TPC on the NuMI beam, a wide program of LArTPCs on neutrino beams has flourished. The construction of  LArTPCs as near and far detectors at different baseline allows for the exploration of some of the fundamental questions in neutrino physics today illustrated in section \ref{ch:questions}.

@@

\subsection{Prospects for LArTPCs in GUT Physics: DUNE}
The experimental exploration of a manifestation of Grand Unified Theory is possible in DUNE thanks to its sheer mass.  In particular, proton decay searches are a capital topic of DUNE's wide non-accelerator physics program.
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  As shown in section \ref{ch:GUTsTheory}, different family of GUTs predict the proton to decay in different modes. In particular, SUSY flavored GUTs prefer the presence of kaons in the decay products, e.g. $p \rightarrow K^+ \bar{\nu}$.
It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold in water.  Thus, Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying  on photons from nuclear de-excitation and on the muon tagging in the kaon decay leptonic mode. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a LArTPCs, where the kaon is directly visible in the detector. 
According to \cite{Adams:2013qkq}, DUNE will have an active volume large enough, have sufficient shielding from the surface, and will run for lengths of time sufficient to compete with Hyper-K, opening up the opportunity for the discovery of nucleon decay.

@@

%\subsection{Non-Accelerator Physics Program}
%\subsection{Rare Decay Searches: Experimental Limit}
%\subsection{Nucleon Decay Detection in LAr}
\subsection{Enabling the next generation of discoveries: LArIAT}
LArIAT, a small LArTPC in a test beam,  is designed to perform an extensive physics campaign centered on charged particle cross section measurements while characterizing the detector performance for future LArTPCs. Since LArTPCs represent the most advanced experiments for physics at the Intensity Frontier, their complex technology needs a thorough calibration and dedicated measurements of some key quantities to achieve the precision required for the next generation of discoveries.  LArIAT's goal is to provide such calibration and dedicated measurements. The LArIAT LArTPC is deployed in a dedicated calibration test beamline at Fermilab. We use the LArIAT beamline to characterize the charge particles before they enter the TPC: the particle type and initial momentum is known from beamline information. The precise calorimetric energy reconstruction of the LArTPC technology enables the measurement of the total differential cross section for  tagged hadrons. 
The Pion-Nucleus and Kaon-Nucleus total hadronic interaction cross section have never been measured before in argon and they are a fundamental step to shed light on light meson interaction in nuclei per se, while providing a key input to neutrino physics and proton decay studies in future LArTPC experiments like SBN and DUNE.

>>>>

\chapter{Liquid Argon Detectors at the Intensity Frontier}\label{ch:2}
{\raggedleft ``\emph{Don't you know, honey,} \par}
{\raggedleft \emph{Ain't nobody ever gonna love you, the way I try to do?}"\par}
{\raggedleft -- Janis Joplin, Cry Baby, 1971 -- \par}
\vspace{0.5cm}

In the next few years, LArTPCs will  be the tools to answer some of the burning questions in neutrino physics today.  This chapter illustrates the operational principles of this detector technology, as well as the scope of the key detectors in the US liquid argon program -- SBN, DUNE and LArIAT.

\section{The Liquid Argon Time Projection Chamber Technology}
In this section, we outline an extremely brief history of Time Projection Chambers as particle detectors, focusing on their incarnation as Argon detectors for neutrino physics. We further describe the working principles of Liquid Argon Time Projection Chambers, leading  to the description of the event reconstruction in LArTPC.

\subsection{TPCs, Neutrinos \& Argon}
David Nygren designed the first Time Projection Chamber (TPC) in the late 1970s~\cite{FirstTPC} for the  PEP-4 experiment, a detector  apt to study electron-positron collisions at the PEP storage ring at the SLAC National Accelerator Laboratory.
From the original design  in the seventies -- a cylindrical chamber filled with methane gas -- the TPC detector concept has seen many incarnations, the employment of several different active media and a variety of different particle physics applications, including, but not limited to the study of electron/positron storage rings (e.g. PEP4, TOPAZ, ALEPH and DELPHI), heavy ions collisions in fixed target and collider experiments (e.g. EOS/HISS and ALICE ), dark matter (ArDM), rare decays and capture (e.g. TRIUMP, MuCap),  neutrino detectors and nucleon decay (ICARUS, SBN, DUNE), and neutrino less double beta decay (Next). A nice review of the history of TPCs and working principles is provided in \cite{0034-4885-73-11-116201}.

@@

\section{The Intensity Frontier Program}
This section highlights the role of Liquid Argon Time Projection Chambers at the Intensity frontier. In particular, we show the prospects for the exploration of neutrino physics (Section \ref{ch:NuPhysLAr}) and GUT models (Section \ref{ch:GutsLAr}) in current and forthcoming LAr experiments. In Section \label{ch:LArIATIntro}, we introduce LArIAT and its role in the Intensity Frontier panorama.

\subsection{Prospects for LArTPCs in Neutrino Physics: SBN and DUNE}\label{ch:NuPhysLAr}
The ArgoNeut  experiment \cite{ArgoNeuT-det} together the LAr R\&D experiments TallBo and the Yale TPC  initiated the US LArTPC neutrino program. Following the success of the ArgoNeut small TPC on the NuMI beam, a wide program of LArTPCs on neutrino beams has flourished. The construction of  LArTPCs as near and far detectors at different baseline allows for the exploration of some of the fundamental questions in neutrino physics today illustrated in section \ref{ch:questions}.

@@

\subsection{Prospects for LArTPCs in GUT Physics: DUNE}\label{ch:GutsLAr}
The experimental exploration of a manifestation of Grand Unified Theory is possible in DUNE thanks to its sheer mass.  In particular, proton decay searches are a capital topic of DUNE's wide non-accelerator physics program.
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  As shown in section \ref{ch:GUTsTheory}, different family of GUTs predict the proton to decay in different modes. In particular, SUSY flavored GUTs prefer the presence of kaons in the decay products, e.g. $p \rightarrow K^+ \bar{\nu}$.
It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold in water.  Thus, Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying  on photons from nuclear de-excitation and on the muon tagging in the kaon decay leptonic mode. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a LArTPCs, where the kaon is directly visible in the detector. 
According to \cite{Adams:2013qkq}, DUNE will have an active volume large enough, have sufficient shielding from the surface, and will run for lengths of time sufficient to compete with Hyper-K, opening up the opportunity for the discovery of nucleon decay.

@@

%\subsection{Non-Accelerator Physics Program}
%\subsection{Rare Decay Searches: Experimental Limit}
%\subsection{Nucleon Decay Detection in LAr}
\subsection{Enabling the next generation of discoveries: LArIAT}\label{ch:LArIATIntro}
LArIAT, a small LArTPC in a test beam,  is designed to perform an extensive physics campaign centered on charged particle cross section measurements while characterizing the detector performance for future LArTPCs. Since LArTPCs represent the most advanced experiments for physics at the Intensity Frontier, their complex technology needs a thorough calibration and dedicated measurements of some key quantities to achieve the precision required for the next generation of discoveries.  LArIAT's goal is to provide such calibration and dedicated measurements. The LArIAT LArTPC is deployed in a dedicated calibration test beamline at Fermilab. We use the LArIAT beamline to characterize the charge particles before they enter the TPC: the particle type and initial momentum is known from beamline information. The precise calorimetric energy reconstruction of the LArTPC technology enables the measurement of the total differential cross section for  tagged hadrons. 
The Pion-Nucleus and Kaon-Nucleus total hadronic interaction cross section have never been measured before in argon and they are a fundamental step to shed light on light meson interaction in nuclei per se, while providing a key input to neutrino physics and proton decay studies in future LArTPC experiments like SBN and DUNE.

### 2018-05-24 Chapter-3/3-TheLArIATExperiment.tex
s

\chapter{LArIAT: Liquid Argon In A Testbeam}\label{sec:experimentDescription}
{\raggedleft ``\emph{But, hey we need to be somewhat foolish...} \par}
{\raggedleft -- Agnes Obel, Avenue, 2010 -- \par}
\vspace{0.5cm}

@@

Events satisfying the simplest scenario of one single hit in each of the four wire chambers form the ``Picky Track" sample.  We construct another, higher statistics sample, where we loosen the requirements on single hit and wire chamber efficiency: the ``High Yield" sample. For LArIAT Run II, the High Yield sample is about three times the Picky Tracks statistics.

@@

\textcolor{red}{Four point track momentum uncertainty}

@@

\section{In the Cryostat}
\subsection{Cryogenics and Argon Purity}\label{ch:Cryo}
LArIAT repurposed the ArgoNeuT cryostat \cite{ArgoNeuT-det} in order to use it in a beam of charged particles, and added a new process piping and a new liquid argon filtration system in FTBF.  %The main modifications to the ArgoNeuT cryostat are the addition of a beam flange and a port for the light collection system, 
Inside the LArIAT experimental hall, the cryostat sits in the beam of charged particles with its horizontal main axis oriented parallel to the secondary beam, ~3$^\circ$ off axis from the tertiary beam

>>>>

\chapter{LArIAT: Liquid Argon In A Testbeam}\label{sec:experimentDescription}
{\raggedleft ``\emph{But, hey we need to be somewhat foolish...}" \par}
{\raggedleft -- Agnes Obel, Avenue, 2010 -- \par}
\vspace{0.5cm}

@@

Events satisfying the simplest scenario of one single hit in each of the four wire chambers form the ``Picky Track" sample.  We construct another, higher statistics sample, where we loosen the requirements on single hit and wire chamber efficiency: the ``High Yield" sample. For LArIAT Run II, the High Yield sample is about three times the Picky Tracks statistics.  
We assume an uncertainty of 2\% for four-point WC track, momentum uncertainty as reported for the same beamline in \cite{MinervaTestbeam}.

@@

@@

\section{In the Cryostat}
The heart of the LArIAT experiment lives in the LArIAT cryostat. In this section, we describe the cryogenic system and the argon purity (Section \ref{ch:Cryo}), the LArIAT TPC (Section \ref{sec:TPCCharge}) and light collection system (\ref{sec:TPCLight}).
\subsection{Cryogenics and Argon Purity}\label{ch:Cryo}
LArIAT repurposed the ArgoNeuT cryostat \cite{ArgoNeuT-det} in order to use it in a beam of charged particles, and added a new process piping and a new liquid argon filtration system in FTBF.  %The main modifications to the ArgoNeuT cryostat are the addition of a beam flange and a port for the light collection system, 
Inside the LArIAT experimental hall, the cryostat sits in the beam of charged particles with its horizontal main axis oriented parallel to the secondary beam, ~3$^\circ$ off axis from the tertiary beam

### 2018-05-24 Chapter-7/7-KaonCrossSection.tex
s

\chapter{Positive Kaon Cross Section Measurement}\label{ch:KaonXS}

@@

\section{Results}\label{ch:FinalKaon}
\section{Results}\label{ch:FinalKaon}
Figure \ref{fig:FinalXSKaon} show the measurement of the ($K^+$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and efficiency correction to the raw cross section. The plot shows the measurement obtained on the full dataset, statistical uncertainty in black and systematic uncertainty in red. The Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

The systematic uncertainty on the cross section is the sum in quadrature of the statistical uncertainty, the systematic uncertainty related to the kinetic energy measurement, the systematic uncertainty related to the beam composition and the systematic uncertainty related to the efficiency correction.

>>>>

\chapter{Positive Kaon Cross Section Measurement}\label{ch:KaonXS}
{\raggedleft ``\emph{Beat-up little seagull, on a marble stair} \par}
{\raggedleft \emph{Tryin' to find the ocean, lookin' everywhere.}"\par}
{\raggedleft -- Nina Simone, Baltimore, 1978 -- \par}
\vspace{0.5cm}

@@

\section{Results}\label{ch:FinalKaon}
Figure \ref{fig:FinalXSKaon} show the measurement of the ($K^+$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and efficiency correction to the raw cross section. The plot shows the measurement obtained on the full dataset, statistical uncertainty in black and systematic uncertainty in red. The Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

The systematic uncertainty on the cross section is the sum in quadrature of the statistical uncertainty, the systematic uncertainty related to the kinetic energy measurement and the systematic uncertainty related to the efficiency correction.
